---
title: "Probability and Simulations - Seminar 2c"
author: "Jennifer Grants"
date: "1/22/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

```{r include=FALSE}
library(tidyverse)
```

# 1. Definitions 

## 1.a. Probability density functions 

Probability density functions map the values of x (measured variable) to associated probability described by the distribution. 

e.g. Function dnorm() 

```{r}
xValues <- seq(from = -5, to = 5, length = 100)
plotA <- dnorm(xValues, mean = 0, sd = 0.5) #
plotB <-  dnorm(xValues, mean = 0, sd = 1)
plotC <-  dnorm(xValues, mean = 0, sd = 2)
plotD <-  dnorm(xValues, mean = -2, sd = 1)

normalDistributionsTibble <- tibble(x_value = xValues, 
                                    red_value = plotA,
                                    blue_value = plotB,
                                    green_value = plotC,
                                    purple_value = plotD)

p <- normalDistributionsTibble %>% ggplot()
p + 
  geom_line(aes(x = xValues, y = red_value), color = "red") +
  geom_line(aes(x = xValues, y = blue_value), color = "blue") +
  geom_line(aes(x = xValues, y = green_value), color = "green") +
  geom_line(aes(x = xValues, y = purple_value), color = "purple") +
  xlab("x") +
  ylab("f(x)")
```


## 1.b. Cumulative Distribution Function (CDF) 

CDF describes the cumulative probability of all values smaller than x. This is why it starts at 0 and goes up to 1, as the probability of all values smaller than x grows with x. The slope corresponds to the spread. 

e.g. Function: pnorm()

